---
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    #theme: flatly
---


<div>
<div style="text-align:center;"> 
  <h1><B>DATA 607 - Assignment 2- Week 2A SQL and R: Code Base submission   </B></h1>
  <h3><B>Bikash Bhowmik ---- 12 Sep 2026</B></h3> 
</div>

</div>


<style>
/* Make the left column fixed (sticky) while scrolling */
.section.level3:first-of-type {
  position: sticky;
  top: 10px;
}

/* Optional: add smooth scroll for anchor links */
html {
  scroll-behavior: smooth;
}

/* Hide the top title/header bar */
.navbar, .navbar-default {
  display: none !important;
}

body {
  font-size: 12pt;
}
</style>



<!-- Include Split.js library -->
<script src="https://unpkg.com/split.js/dist/split.min.js"></script>

<div id="split-container" style="height: 100vh; display: flex; flex-direction: row;">

  <!-- Left column -->
  <div id="left-panel" style="width:150px; overflow:auto; background:#f8f9fa; padding:10px;">

### 
<a href="#mySection1" style="font-size: 1em; font-weight: bold;" >Objective</a>
<br>
<a href="#mySection2" style="font-size: 1em; font-weight: bold;" >Data Collection</a>
<br>
<a href="#mySection2a" style="font-size: 1em; font-weight: bold;" >Tools / Environment</a>
<br>
<a href="#mySection2b" style="font-size: 1em; font-weight: bold;" >Database Design</a>
<br>
<a href="#mySection2bb" style="font-size: 1em; font-weight: bold;" >Keys and Constraints (Planned)</a>
<br>
<a href="#mySection2c" style="font-size: 1em; font-weight: bold;" >Missing Data Strategy (Planned)</a>
<br>
<a href="#mySection3" style="font-size: 1em; font-weight: bold;" >Analysis Plan in R</a>
<br>
<a href="#mySection3a" style="font-size: 1em; font-weight: bold;" >Reproducibility Plan</a>
<br>
<a href="#mySection4" style="font-size: 1em; font-weight: bold;" >Loading Libraries</a>
<br>
<a href="#mySection41" style="font-size: 1em; font-weight: bold;" >Connect to PosgreSQL</a>
<br>
<a href="#mySection5" style="font-size: 1em; font-weight: bold;" >Load table in R</a>
<br>
<a href="#mySection5_0" style="font-size: 1em; font-weight: bold;" >Join tables in R</a>
<br>
<a href="#mySection50" style="font-size: 1em; font-weight: bold;" >Demonstrate missing ratings handling (Part 1)</a>
<br>
<a href="#mySection500" style="font-size: 1em; font-weight: bold;" >Demonstrate missing ratings handling (Part 2)</a>
<br>
<a href="#mySection51" style="font-size: 1em; font-weight: bold;" >Simple summaries</a>
<br>
<a href="#mySection6" style="font-size: 1em;font-weight: bold; " >Disconnecting Database</a>
<br>
<a href="#mySection61" style="font-size: 1em;font-weight: bold; " >Conclusion</a>
<br>
 </div>
 
  <!-- Right column -->
  <div id="right-panel" style="flex-grow:1; overflow:auto; padding:10px;">
  
<a id="mySection1"></a>
<font size="4">
<B>
Objective
</B>
</font>

The goal of this assignment is to collect a small dataset of movie ratings from different users, store the data in a PostgreSQL database, and analyze it using R. This project demonstrates the complete process of collecting data, storing it in a relational database, querying the data, and bringing it into R for analysis. It also shows how missing movie ratings can be handled appropriately.

<a id="mySection2"></a>
<font size="4">
<B>
Data Collection Plan
</B>
</font>


-   **Items (Movies):** I will select **six recent popular movies** (or similar media items) to increase the chance that participants have seen at least some of them.
-   **Users (Participants):** I will survey **at least five people**.
-   **Rating Scale:** Participants will rate any movie they have seen on a **1–5 integer scale**.
-   **Missing Ratings:** If a participant has not seen a movie, the rating will be recorded as **missing (NULL)** instead of forcing a guess.

Responses will be collected in a simple format (e.g., a small table) and then inserted into PostgreSQL.

<a id="mySection2a"></a>
<font size="4">
<B>
Tools / Environment
</B>
</font>


-   **Database:** PostgreSQL (local installation)\
-   **GUI:** pgAdmin 4 (for running SQL scripts and verifying tables)\
-   **R Packages (planned):** `DBI`, `RPostgres`, `tidyverse`

> **Security Note:** Database passwords will **not** be included in the code. Connection credentials will be stored using environment variables or masked placeholders.


<a id="mySection2b"></a>
<font size="4">
<B>
Database Design (Normalized Schema)
</B>
</font>

To keep the solution professional and interview-ready, I will use a **normalized relational schema** to represent the many-to-many relationship between users and movies.

Planned tables:

-   `users(user_id, name)`
-   `movies(movie_id, title, release_year)`
-   `ratings(rating_id, user_id, movie_id, rating)`
`

<a id="mySection2bb"></a>
<font size="4">
<B>
Keys and Constraints (Planned)
</B>
</font>


-   `users.user_id` and `movies.movie_id` will be **primary keys**
-   `ratings.user_id` and `ratings.movie_id` will be **foreign keys**
-   Ratings will be constrained to the range **1–5**
-   A **unique constraint** on `(user_id, movie_id)` will prevent duplicate ratings

This structure mirrors how real recommendation datasets are stored: users and movies are separate entities, and ratings are stored in a junction table.


<a id="mySection2c"></a>
<font size="4">
<B>
Missing Data Strategy
</B>
</font>


Missing ratings are expected because participants may not have seen every movie.

I will handle missing data in two ways:

1.  **In SQL:** Missing ratings will be represented as `NULL` (or omitted rows if a participant did not rate an item).
2.  **In R:** When analyzing results, missing ratings will appear as `NA`. Summary statistics (means, medians) will be computed with `na.rm = TRUE` to avoid bias from missingness.

To document missingness, I will report:

-   ratings count per movie (coverage)
-   ratings count per user
-   how many missing ratings exist overall

<a id="mySection3"></a>
<font size="4">
<B>
Analysis Plan in R
</B>
</font>

After the database is populated, I will connect PostgreSQL to R using DBI and RPostgres. I will then use R to:

-Combine the users, movies, and ratings tables into one tidy dataset.<br>
-Calculate basic summaries, including:<br>
      -The average rating for each movie along with the number of ratings.<br>
      -The average rating given by each user.<br>
      -The overall distribution of ratings from 1 to 5.<br>
-Optionally, convert the data into a user–item matrix in wide format to show how the dataset can be structured for collaborative filtering and recommendation systems. An advanced recommendation model is not required for this assignment.

<a id="mySection3a"></a>
<font size="4">
<B>
Reproducibility Plan
</B>
</font>


Even if I use pgAdmin to run queries, I will include the full SQL scripts required to:

-   create the tables (`CREATE TABLE`)
-   populate them (`INSERT INTO`)
-   query them (`SELECT ... JOIN ...`)

All code (SQL + R + Quarto) will be stored in a GitHub repository for submission, with sensitive credentials removed or masked.

<a id="mySection4"></a>
<font size="4">
<B>
Loading Libraries
</B>
</font>

library(DBI) <br>
library(RPostgres)<br>
library(dplyr)<br>
library(tidyr)<br>


```{r global_options}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
library(DBI)
library(RPostgres)
library(dplyr)
library(tidyr)

```


<a id="mySection41"></a>
<font size="4">
<B>
Connect to PosgreSQL
</B>
</font>

```{r}
con <- dbConnect(
  RPostgres::Postgres(),
  host = "localhost",
  port = 5432,
  dbname = "postgres",
  user = "postgres",
  password = "admin"
)

dbListTables(con) 

```

<a id="mySection50"></a>
<font size="4">
<B>
Load table in R
</B>
</font>
```{r}
users_df  <- dbGetQuery(con, "SELECT * FROM users;")
movies_df <- dbGetQuery(con, "SELECT * FROM movies;")
ratings_df <- dbGetQuery(con, "SELECT * FROM ratings;")

```

<a id="mySection5_0"></a>
<font size="4">
<B>
Join tables in R
</B>
</font>


```{r}
ratings_joined <- ratings_df %>%
  left_join(users_df, by = "user_id") %>%
  left_join(movies_df, by = "movie_id") %>%
  select(name, title, release_year, rating) %>%
  arrange(name, title)

ratings_joined
```


<a id="mySection50"></a>
<font size="4">
<B>
Demonstrate missing ratings handling (Part 1)
</B>
</font>


```{r}
ratings_complete <- users_df %>%
  crossing(movies_df) %>%
  left_join(ratings_df, by = c("user_id", "movie_id"))

ratings_complete


```

<a id="mySection500"></a>
<font size="4">
<B>
Demonstrate missing ratings handling (Part 2)
</B>
</font>

```{r}

ratings_complete %>%
  summarise(
    total_possible = n(),
    missing_ratings = sum(is.na(rating)),
    observed_ratings = sum(!is.na(rating))
  )


```


<a id="mySection51"></a>
<font size="4">
<B>
Simple summaries
</B>
</font> 

```{r}
ratings_complete %>%
  group_by(title) %>%
  summarise(avg_rating = mean(rating, na.rm = TRUE),
            n_ratings = sum(!is.na(rating)))
```



<a id="mySection6"></a>
<font size="4">
<B>
Disconnecting Database
</B>
</font>

```{r}
dbDisconnect(con)
```


<a id="mySection61"></a>
<font size="4">
<B>
Conclusion
</B>
</font>

This assignment demonstrates how relational databases and R can be used together in a data science workflow. PostgreSQL provides structured data storage and helps maintain data integrity through constraints, while R is used for data manipulation, handling missing values, and performing analysis. The final dataset is organized in a user–item ratings format, which is commonly used in recommendation systems.

 </div>

</div>



<style>
/* Make the left column fixed (sticky) while scrolling */
.section.level3:first-of-type {
  position: sticky;
  top: 10px;
}

/* Optional: add smooth scroll for anchor links */
html {
  scroll-behavior: smooth;
}

/* Hide the top title/header bar */
.navbar, .navbar-default {
  display: none !important;
}


</style>

<script>
Split(['#left-panel', '#right-panel'], {
  sizes: [20, 80],      // initial width percentage
  minSize: [150, 300],  // min width in pixels
  gutterSize: 6,        // width of the draggable gutter
  cursor: 'col-resize'
});
</script>

<style>
.gutter {
  background-color: #ddd;
  cursor: col-resize;
}
</style>





